---
title: "Boxplots"
output: 
  learnr::tutorial:
    progressive: true
    theme: cerulean
    highlight: pygments
    css: css/test.css
    code_folding: hide
runtime: shiny_prerendered
author: Rob Knell
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(echo = TRUE)
WorldBank <- TestTutorial::WorldBank

WorldBank$Region <- as.factor(WorldBank$Region)
```


## What is a boxplot?

A boxplot is a way of visualising data which displays a lot of information while still being simple and easy to interpret. They are particularly useful when there you have multiple variables or different factor levels associated with a numeric variable and you want to compare them. As an example, let's have a look at the data on population growth from the WorldBank dataset. We'll start by looking at a frequency histogram.

```{r fig.cap = "Frequency histogram of annual percentage population growth for 186 countries", fig.height = 6}

hist(WorldBank$Population_growth, 
     main = "",
     col = "aquamarine4",
     xlab = "Annual population growth per country (%)")
```


You can see that this variable is roughly normally distributed, with a central tendency somewhere between 1 & 2 and a range from just below -4 to somewhere between 6 & 7. We can get more exact numbers using `summary()`

```{r}
summary(WorldBank$Population_growth)
```

Rather than looking at a frequency histogram, we can use a boxplot to give us much the same information. The function to do this in R is conveniently called `boxplot()`. 

```{r fig.cap = "Boxplot of annual population growth for 186 countries", fig.height = 6}
boxplot(WorldBank$Population_growth,
        col = "aquamarine4",
        ylab = "Annual population growth per country (%)")
```
If you've not seen one of these before it might be a bit confusing so let's go through all the different bits one at a time.

1. There's a thick line in the middle, sometimes called the *hinge*. This indicates the *median*: the middle value when the data are ranked.
2. There's a *box* around the thick line. This extends from the *First quartile* to the *Third quartile*: so this is showing you the range of values within which the middle 50% of all the datapoints lie. NB the width of the box has no meaning, at least for a normal boxplot. 
3. There are dotted lines extending some distance above and below the box. These are called the *whiskers* and they extend wait for it from the quartile (so the end of the box) to the data point which is nearest to the last datapoint which is less than 1.5 times the *interquartile range* from the box. This might sound a little random but there is a good reason behind it: if the data are drawn from a normal distribution, roughly 99% (actually 99.3%) of all the data should fall within this range. NB you will sometimes see boxplots drawn with the whiskers extending up to the maximum and down to the minimum values. There's nothing especially wrong with doing this but it does reduce the amount of information in the graph.
4.  Finally, any datapoints which lie outside the whiskers are plotted individually. These are usually called *outliers* but be careful with this word. Recall that we'd expect about 99% of the data to be within the whiskers if the data are from a normal distribution, so with 100 datapoints we should expect at to see one or more of these "outliers" most of the time. Here we have 186 datapoints and three "outliers" and that is entirely unsurprising. There's no reason to think that there is anything strange about these datapoints, and there is no reason to think that they are perhaps from a differently distributed set of data to the rest. It's better to think of them as *extreme values* rather than *outliers* because the latter is often used for datapoints that are somehow in the wrong dataset.

Overall then a boxplot shows you the median, the interquartile range, the region within which roughly 99% of datapoints would be expected if the underlying distribution is normal, and any datapoints outside this region. To make this clearer, here is our frequency histogram again, this time with the boxplot plotted above it.

```{r fig.height = 7, fig.width = 7, echo = TRUE}
# Set plot area for histogram to the lower 70% of the total area
par(fig = c(0, 1, 0, 0.7))

# Plot the frequency histogram
hist(
  WorldBank$Population_growth,
  main = "",
  xlab = "Population growth (%)",
  breaks = 15,
  col = "aquamarine4"
)


# set the plot area for the boxplot to the upper 60% of the total area
par(fig = c(0, 1, 0.4, 1), new = TRUE)

# Plot the boxplot
boxplot(WorldBank$Population_growth,
        col = "aquamarine4",
        bg = FALSE,
        horizontal = TRUE,
        axes = FALSE)
```

You can see how the median and the interquartile range indicated on the boxplot correspond to the centre of the distribution and the region with the bulk of the data present, and you can also see how the whiskers cover almost all of the range of the data, with the few extreme values showing up individually. 

### Exercise: draw a boxplot of CO~2~ production by country

There is a variable in the WorldBank dataset called CO2 which is the annual CO~2~ production of each country in tonnes per capita. See if you can draw a boxplot of this variable and label the y-axis "Annual Carbon Dioxide Production Per Capita (t)". 

```{r boxplot1, exercise = TRUE, fig.height = 6}

```

```{r boxplot1-hint-1}
#Remember that you need to specify the WorldBank 
#dataframe and give the name of the CO2 variable 
#with the two separated by a dollar symbol
```

```{r boxplot1-hint-2}
#Use the ylab = argument to specify the y-axis label. 
#The text for the label needs to be in quote marks.
```

```{r boxplot1-hint-3}
#Check that there's a comma between all arguments 
#and that all your brackets and quote marks are 
#matched.
```

```{r boxplot1-hint-4}
boxplot(WorldBank$CO2,
ylab = "Annual Carbon Dioxide Production Per Capita (t)")
```

This boxplot looks very different to the one we drew in the last section. It looks as though it's been squashed towards the bottom and stretched towards the top, and all of the extreme values that are shown are relatively large values. To understand what we're seeing here, it will help to plot a frequency histogram for CO~2~ production. Don't forget to label the axes.

```{r boxplot2, exercise = TRUE, fig.height = 6}

```

```{r boxplot2-hint-1}
# You need to use the hist() function
# The y-axis label should just be "Frequency"
# The x-axis needs a label which says what it is
# Remember that the text for axis labels goes in quote marks
```

```{r boxplot2-hint-2}
#Check that there's a comma between all arguments 
#and that all your brackets and quote marks are 
#matched.
```

```{r boxplot2-hint-3}
hist(WorldBank$CO2,
xlab = "Annual Carbon Dioxide Production Per Capita (t)",
ylab = "Frequency", 
main = "")
```

You can see from looking at the frequency histogram that unlike the population growth data, per capita CO~2~ production is strongly positively skewed. This accounts for the different shapes of the two boxplots. The approximately normal distribution of the population growth data gives us a boxplot which is roughly symmetrical above and below the median, but the strong positive skew in the CO~2~ production data gives a very asymmetrical boxplot, with the lower whisker, quartile and the median being close together and the rest of the plot looking as though it's been stretched upwards. This shows us one of the great strengths of using boxplots as part of your initial exploration of your data: they don't just give you information on where the data are located, they tell you about the shape of the data as well.

As a final exercise, here is the code we used before to draw the frequency histogram for population growth with the boxplot above it. See if you can modify this to draw the equivalent graph for CO~2~ production. Try changing the colour from "aquamarine4" to "steelblue" and don't forget to make sure the axis labels are correct.

```{r boxplot2a, exercise = TRUE, fig.height = 7, fig.width = 7}
par(fig = c(0, 1, 0, 0.7))
hist(
  WorldBank$Population_growth,
  main = "",
  xlab = "Population growth (%)",
  breaks = 15,
  col = "aquamarine4"
)

par(fig = c(0, 1, 0.45, 1), new = TRUE)

boxplot(WorldBank$Population_growth,
        col = "aquamarine4",
        horizontal = TRUE,
        axes = FALSE)
```

## Multiple boxplots

Boxplots can show you the shape and the location of your data, but so can frequency histograms. Why don't we just plot histograms? The answer to this is that boxplots become really useful when we have multiple groups of data that we want to compare. If you have a variable and also a factor which divides the data in your variable into groups, then `boxplot()` will allow you to visualise this. Instead of entering a single variable name you need to enter a *formula*, with the variable on the left and the factor name on the right, with a tilde between: `boxplot(variable ~ factor)`.

As an example, here are our population growth data plotted as boxplots with the different regions in our WorldBank dataset plotted separately. There are a few things that you might not be familiar with:

`par(mar = c(12,4,2,2))` sets the margins for the plot, in the order bottom, left, top, right. I've made the bottom margin much bigger than usual because the region names are long and need plenty room. I've also added the `las = 2` argument in the `boxplot()` function call. This makes the text for the axis labels perpendicular to the axis, so the x-axis labels are vertical.

```{r fig.cap = "Population growth data plotted for each region from the WorldBank dataset", fig.height = 6}

par(mar = c(12,4,2,2))

boxplot(Population_growth ~ Region,
        data = WorldBank,
        ylab = "Annual population growth (%)",
        xlab = "",
        las = 2)
```

Now you can see how a plot like this can convey a huge amount of information. You can:

* Find out about the central tendencies of each group, so the Europe & Central Asia group has the lowest median growth rate. 
* Make comparisons between groups regarding how much variability there is: so North America hs the lowest variability (but also only three countries), Latin America & the Caribbean has fairly low variability but the Middle East and North Africa region has the most variability. 
* Use it to check the likely error distribution: here you can see that most of the boxplots are at least roughly symmetrical, so there's probably going to be no problem with using "standard" analysis techniques that assume normal errors.
* Use it to check for potentially problematic datapoints --- that one data point with a growth rate of < -4% in the Middle East & North Africa group seems a little odd and you might want to check it...

```{r}
WorldBank[WorldBank$Population_growth < -4,]
```

Oh, it's Syria. I guess the declining population makes sense then. Not the happiest thought.

### Exercise: draw a boxplot of CO~2~ production by Income Group

The WorldBank dataset also includes a variable called `Income_group` which splits the countries into four classes: Low income, Lower middle income, Upper middle income and High income. Let's generate a boxplot which compares the CO~2~ production data for these four groups. 

Before we draw the boxplot we'll have to declare `Income_group` as a factor and just to make the graph make more sense we'll change the order of the factor levels: R orders factor levels alphabetically by default, but that would generate a plot that's not so easy to interpret. The following piece of code does this for us:

```{r eval = FALSE, echo = TRUE}
WorldBank$Income_group <- factor(
  WorldBank$Income_group,
  levels = c(
    "Low income",
    "Lower middle income",
    "Upper middle income",
    "High income"
  )
)
```


```{r prepare-boxplot-3, echo = TRUE}
WorldBank$Income_group <- factor(
  WorldBank$Income_group,
  levels = c(
    "Low income",
    "Lower middle income",
    "Upper middle income",
    "High income"
  )
)
```


Now that we've sorted out our factor levels, we just need the code for the boxplot. Remember what we did for the last plot with changing the margins with `par(mar = ...` and using the `las=` argument in the `boxplot()` call to change the angle that the axis labels were written at.

```{r boxplot3, exercise = TRUE, exercise.lines = 9, exercise.setup = "prepare-boxplot-3", fig.height = 6, fig.cap = "Annual CO~2~ Production in tonnes per capita plotted by income group"}


```

```{r boxplot3-hint-1}
# You can use the same code we used for the previous example.
# You just need to change the two variable names and the axis labels.
```

```{r boxplot3-hint-2}
#Check that there's a comma between all arguments 
#and that all your brackets and quote marks are 
#matched.
```

```{r boxplot3-hint-3}
#This is the solution

par(mar = c(12,4,2,2))

boxplot(CO2 ~ Income_group, 
data = WorldBank,
xlab = "",
ylab = "Per capita CO2 production (tonnes)",
las = 2
)
```

Once again this boxplot gives us a great deal of information. You can clearly see the strong relationship between income group and per capita CO~2~ production, and you can also see from the asymmetric shape of the boxplots that the frequency distribution is somewhat positively skewed within each group, although not as strongly as when we plotted the whole variable by itself in the last section. This is something that we would need to bear in mind if we wished to analyse these data using, for example, ANOVA, which assumes that the error distribution within each group is normal. A further concern would arise from the increase in variance as the median gets larger --- you can see that the overall variation within the low income group is very small by comparison with the high income group, and again this *heteroscedasticity* would be a concern if we wished to analyse these data using ANOVA.

### Exercise: Log-transforming the y-axis

One possible solution to the skewed errors and heteroscedasticity in this dataset would be to log transform the data prior to analysis. Before doing this it would be a good idea to plot the data on a log scale to make sure that the transformation is making the data  behave as we'd like. There are several ways to do this in R: we could just log transform the variable, either before plotting it:

`logCO2 <- log10(WorldBank$CO2`
`boxplot(logCO2 ~ WorldBank$Income-group...`

or within the `boxplot()` function call:

`boxplot(log(CO2) ~ Income_group, data = WorldBank, ...`.

An alternative is to transform the *scale* rather than the data. This converts the y-axis (in this case) to a log scale and then the untransformed data are plotted. For purposes of visualisation this often works better because the scale retains the real values of the data. We can ask R to plot the data in this way by adding another argument to the `boxplot()` function call, `log = "y"`. Note that if we were plotting a scatterplot then we could plot a log-scale x-axis instead with `log = "x"` or we could have both axes on a log scale with `log = "xy"`. Because the only continuous axis in our boxplot is the y-axis it only makes sense to change the scale of that one.

Try to plot your boxplot with a log-scaled y-axis.

```{r boxplot4, exercise = TRUE, exercise.lines = 9, exercise.setup = "prepare-boxplot-3", fig.height = 6, fig.cap = "Log annual CO~2~ Production in tonnes per capita plotted by income group"}

```


```{r boxplot4-hint-1}
# You can use the same  code as before
# You just need to add the new argument
```


```{r boxplot4-hint-2}
#Check that there's a comma between all arguments 
#and that all your brackets and quote marks are 
#matched.
```

```{r boxplot4-hint-3}
#This is the solution

par(mar = c(12,4,2,2))

boxplot(CO2 ~ Income_group, 
data = WorldBank,
xlab = "",
ylab = "Log scaled per capita CO2 production (tonnes)",
las = 2,
log = "y"
)
```

Plotting the data on a log scale gives us a very different graph indeed. The variances are roughly equivalent between the groups and there is only a hint of asymmetry, suggesting that the skew that was present before has been largely dealt with. If you were to wish to compare the mean CO~2~ production figures between these groups using ANOVA there's nothing to indicate potential problems from heteroscedasticity or skewed error distributions here.

## Boxplots with more than one factor

`boxplot()` can plot more complex relations between variables than the examples we've seen. For example, we might be interested in how the amount of forested area is related to income, and might also think that this might vary with latitude, so countries nearer the poles might be different from countries nearer the equator. 

Rather than plotting latitude as a continuous variable we can generate a new factor where we split countries into groups on the basis of the latitude of their capital: here we'll use "Tropical" to mean within < 35 degrees of latitude (this is actually Tropical and Subtropical but let's not make things too complicated) and "Temperate or Polar" to refer to higher latitudes. We'll also set up a new factor which divides countries more simply into high and low income.


```{r, prepare-boxplot-5, echo = TRUE}

Climate <-
  as.factor(ifelse(
    abs(WorldBank$Capital_lat) < 35,
    "Tropical",
    "Temperate or Polar"
  ))

Income <-
  as.factor(
    ifelse(
      WorldBank$Income_group == "High income" |
        WorldBank$Income_group == "Upper middle income",
      "High",
      "Low"
    )
  )
```

OK, let's see what a boxplot of forest area conditioned on climate and income looks like. To use multiple factors in a `boxplot()` call you use `interaction()` within the formula specifying the variables to be drawn.

```{r}

par(mar = c(12,4,2,2))

boxplot(WorldBank$Forest_area. ~ interaction(Income: Climate),
        las = 2,
        ylab = "Percentage of area forested",
        xlab = "")
```

That gives us a nice plot but the x-axis is a bit of a disaster. In some cases it's best to roll your own. Here we've set the `xaxt = "n"` argument within the `boxplot()` function call which stops the x-axis being drawn, and then we're using the `axis()` function to draw our own axis. The `\n` within the labels are *escape characters* which tell R to put a new line there, and `padj = 0.5` just moves the text down a little.

```{r fig.height = 6}
boxplot(WorldBank$Forest_area. ~ interaction(Income: Climate),
        las = 2,
        ylab = "Percentage of area forested",
        xlab = "",
        xaxt = "n")

axis(side= 1,
     at = 1:4,
     padj = 0.5,
     labels = c("High income\nTemperate", "High income\nTropical", "Low income\nTemperate", "Low income\nTropical"))
```

As an alternative to giving al the information in the axis labels, we could colour code the graph. Here's an example:

```{r fig.height = 6, fig.width = 8}

boxplot(WorldBank$Forest_area. ~ interaction(Income: Climate),
        las = 2,
        ylab = "Percentage of area forested",
        xlab = "",
        xaxt = "n",
        col = c("darkgreen", "orange"))

legend("topleft", legend = c("Temperate or Polar", "Tropical"), fill = c("darkgreen", "orange"))

axis(side = 1, at =c(1.5, 3.5), labels = c("High income", "Low income"))

```

